Fixing Vulnerabilities Hinders Maintainability

### 2019-02-03 paper/ICPC19/icpc19.tex
fixing commits; safe --> secure

In 2011, the International Organization for Standardization (ISO) issued an
update for software product quality ISO/IEC 25010~\cite{iso:2011} considering
\emph{Security} as one of the main software product quality characteristics.
Unmaintainable code is hard to test, analyse and re-use, and thus more difficult
to find vulnerabilities and defects. Therefore, writing maintainable code
prevents the introduction of vulnerabilities in the codebases. In this work, we are
concerned about the impact that patching vulnerabilities may have in the
open-source software maintainability. Are developers affecting software
maintainability when trying to improve software safety? We suspect that some of
the patches may have a negative impact in the software maintainability and
possibly even be the cause of the introduction of new vulnerabilities.

ISO describes software maintainability as ``the degree of effectiveness and
efficiency with which a software product or system can be modified to improve
it, correct it or adapt it to changes in environment, and in requirements'' on
software quality ISO/IEC 25010. As ISO does not provide any specific guidelines or
formula to calculate maintainability, we resort on Software Improvement
Group (SIG)'s web-based source code analysis service Better Code Hub
(BCH)~\cite{Visser:2016:OREILLY} to compute the maintainability of a given
project. There are other well-known standards and models that have been proposed to increase software safety:
Common Criteria~\cite{common:2009} which suffered a lot of criticism regarding the
costs associated and poor technical evaluation; the OWASP
Application Security Verification Criteria~\cite{oswap:2009} which is focused only
on web applications, and a model proposed by Xu et al. (2013)~\cite{6616351} for
raiting software security which we believe it was one of the first steps taken
by SIG to introduce security on their maintainability model.

@@

\textbf{\textit{Security refactorings harm the maintainability of open-source software.}}
%
We found that 40.2\% of the security refactoring have a negative impact on the
maintainability of open-source software, Figure~\ref{fig:secvsreg}. This raises
a new tradeoff when developers need to fix vulnerabilities on their projects
because they might be affecting the software maintainability or even introducing
new vulnerabilities. This study exhibits evidence that developers may have to
reduce maintainability for the sake of safety. We understand that developers
should be able to produce secure code without affecting the maintainability of
their projects. But there are still some concerns that need to be addressed:
\begin{itemize}
	\item \textbf{Security Frameworks} need to be secure and user-friendly~\cite{5287006, 7676144}. Unfortunately,
	there are several vulnerabilities that stay undisclosed. Thus, developers should not
	trust them blindly and check defaults.
	\item \textbf{Documentation} of software should provide developers with the best practices to implement safe and maintainable software. Unmaintainable code is difficult to test, analyze and re-use.
	\item \textbf{Proper Authorization Mechanisms} should be designed and implemented. This is
	one of the most complex security features to implement. Some mechanisms might assist authorization maintenance in long run, such as the combination of \texttt{HTTP} filters with \texttt{Spring Security} annotations.
	\item \textbf{Deprecated code} should be avoided by developers. They should integrate tools such as BCH on their projects CI/CD pipelines to help them be more aware of the maintainability of their code and issues (e.g., dead and duplicated code).
	\item\textbf{Programming Languages} should provide new mechanisms to easily implement security patterns without endangering maintainability.

@@

\Luis{A primeira threat que falas é que os resultados podem nao generalizar
para software fechado. Eu acho que deviamos ser mais diretos.}
\Sof{Melhor?}

We use a dataset that only comprises refactorings of open-source software. However, the methodology requires to access data that is not publicly available. Thus, our findings may not extent to non-open source software.

Different programming languages may require different coding practices to
address software safety. The dataset comprises more commits in C and PHP, i.e.,
the dataset may not be properly representative of the population regarding to programming languages.

@@

\Luis{eu nao acho que usar 10 guidelines proposed by the Software Improvement
Group (SIG) seja algo que nos esteja a diferenciar. Acho que o a prinicpal
diferenca é que estamos a estudar maintainability issues noutro contexto.}\Sof{Melhor?}

@@

Other studies investigated the effect of programming languages in software quality~\cite{Ray:2017:LSP:3144574.3126905,
2019arXiv190110220B}. The first study shows that some programming languages are more buggy than others, i.e.,
language design has impact on software quality. However, the authors of the second
study were not able to reproduce the first one and did not obtain any evidence about the
impact of language design. Both have used the number of defects as an indicator of
software quality. The present work studies how security refactorings affect software quality
based on the code maintainability analysis and provides some evidence that the programming language has impact on maintainability.

\Luis{ tambem fizemos analise por linguagem, nesta ultima citacao deviamos falar sobre isso tb.}\Sof{Acho que esta!}

@@

This work presents an empirical study on the impact of 607 security refactorings on
the maintainability of 94 open-source projects. We leveraged Better Code Hub reports
of each project version to calculate maintainability based on a model proposed in
previous work. Security refactorings significantly affect the software maintainability of open-source
software, i.e., developers hinder maintainability when fixing vulnerabilities. We found
this evidence in 40.2\% of the studied cases. Regarding regular commits, it seems that
overall maintainability increases (50.16\%). However, there is still a very considerable amout of
cases that harm maintaniability (41.35\%).

\Luis{secure security frameworks, nao percebi bem o que se quer dizer aí.}
\Sof{Melhor?}

This study also exhibits evidence that some patterns need more attention than
others. \emph{Broken Autentication} issues, \emph{Denial-of-Service} attacks and \emph{Memory Leaks}
endure maintainability in 55.3\%, 51.3\% and 47.5\%, respectively. Following
the implications of our work, we draw attention for the need of trustfull security
frameworks, code best practices documentation to produce safe and maintainable
code, proper authorization mechanisms design and implementation, better tools
to introduce in the CI/CD pipeline to help software maintenance and new or
improved programming languages to allow developers to easily implement safety
in their software without endangering maintainability.

>>>>

In 2011, the International Organization for Standardization (ISO) issued an
update for software product quality ISO/IEC 25010~\cite{iso:2011} considering
\emph{Security} as one of the main software product quality characteristics.
Unmaintainable code is hard to test, analyse and re-use, and thus more difficult
to find vulnerabilities and defects. Therefore, writing maintainable code
prevents the introduction of vulnerabilities in the codebases. In this work, we are
concerned about the impact that patching vulnerabilities may have in the
open-source software maintainability. Are developers affecting software
maintainability when trying to improve software security? We suspect that some of
the patches may have a negative impact in the software maintainability and
possibly even be the cause of the introduction of new vulnerabilities.

ISO describes software maintainability as ``the degree of effectiveness and
efficiency with which a software product or system can be modified to improve
it, correct it or adapt it to changes in environment, and in requirements'' on
software quality ISO/IEC 25010. As ISO does not provide any specific guidelines or
formula to calculate maintainability, we resort on Software Improvement
Group (SIG)'s web-based source code analysis service Better Code Hub
(BCH)~\cite{Visser:2016:OREILLY} to compute the maintainability of a given
project. There are other well-known standards and models that have been proposed to increase software security:
Common Criteria~\cite{common:2009} which suffered a lot of criticism regarding the
costs associated and poor technical evaluation; the OWASP
Application Security Verification Criteria~\cite{oswap:2009} which is focused only
on web applications, and a model proposed by Xu et al. (2013)~\cite{6616351} for
raiting software security which we believe it was one of the first steps taken
by SIG to introduce security on their maintainability model.

@@

\textbf{\textit{Security refactorings harm the maintainability of open-source software.}}
%
We found that 40.2\% of the security refactoring have a negative impact on the
maintainability of open-source software, Figure~\ref{fig:secvsreg}. This raises
a new tradeoff when developers need to fix vulnerabilities on their projects
because they might be affecting the software maintainability or even introducing
new vulnerabilities. This study exhibits evidence that developers may have to
reduce maintainability for the sake of security. We understand that developers
should be able to produce secure code without affecting the maintainability of
their projects. But there are still some concerns that need to be addressed:
\begin{itemize}
	\item \textbf{Security Frameworks} need to be secure and user-friendly~\cite{5287006, 7676144}. Unfortunately,
	there are several vulnerabilities that stay undisclosed. Thus, developers should not
	trust them blindly and check defaults.
	\item \textbf{Documentation} of software should provide developers with the best practices to implement secure and maintainable software. Unmaintainable code is difficult to test, analyze and re-use.
	\item \textbf{Proper Authorization Mechanisms} should be designed and implemented. This is
	one of the most complex security features to implement. Some mechanisms might assist authorization maintenance in long run, such as the combination of \texttt{HTTP} filters with \texttt{Spring Security} annotations.
	\item \textbf{Deprecated code} should be avoided by developers. They should integrate tools such as BCH on their projects CI/CD pipelines to help them be more aware of the maintainability of their code and issues (e.g., dead and duplicated code).
	\item\textbf{Programming Languages} should provide new mechanisms to easily implement security patterns without endangering maintainability.

@@

We use a dataset that only comprises refactorings of open-source software. As
such, our findings may not generalize to non-open source software.

Different programming languages may require different coding practices to
address software security. The dataset comprises more commits in C and PHP, i.e.,
the dataset may not be properly representative of the population regarding to
programming languages.

@@

@@

Other studies investigated the effect of programming languages in software
quality~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}. The first
study shows that there are programming languages that are more prone to bugs
than others. However, the authors of the second study~\cite{2019arXiv190110220B}
were not able to reproduce the first one and did not obtain any evidence about
the impact of language design. Both have used the number of defects as an
indicator of software quality. The present work studies how security
refactorings affect software quality based on the code maintainability analysis
and provides some evidence that the programming language has impact on
maintainability.

@@

This work presents an empirical study on the impact of 607 security refactorings
on the maintainability of 94 open-source projects. We leveraged Better Code Hub
reports of each project version to calculate maintainability based on a model
proposed in previous work. Security refactorings significantly affect the
software maintainability of open-source software, i.e., developers hinder
maintainability when fixing vulnerabilities. We found this evidence in 40.2\% of
the studied cases. Regarding regular commits, we observed that the overall
maintainability increases (50.16\%). However, there is still a very considerable
amout of cases that harm maintaniability (41.35\%).

Our study also shows evidence that there are patterns that need more attention
than others. \emph{Broken Autentication}, \emph{Denial-of-Service}, and
\emph{Memory Leaks} endure maintainability in 55.3\%, 51.3\% and 47.5\%,
respectively. Following the implications of our work, we draw attention for the
need of code best practices  documentation to produce secure and maintainable
code, proper authorization mechanisms design and implementation, better tools to
introduce in the CI/CD pipeline to help software maintenance and new or improved
programming languages to allow developers to easily implement security in their
software without endangering maintainability.
